---
title: "Software"
description: |
  Open source software.
output: 
  distill::distill_article:
    css: quang.css
---

```{r setup, include=FALSE}
library(htmltools)
source("R/html_utils.R")
knitr::opts_chunk$set(echo = FALSE)
```

## Research Software  
```{r}
create_card(
  url = "https://github.com/qpmnguyen/CBEA", 
  title = "CBEA: Competitive balances for taxonomic enrichment analysis",
  source = "https://github.com/qpmnguyen/CBEA",
  description = "R package for performing enrichment analysis using the CBEA approach. Contains additional support function for generalized set enrichment analysis for microbiome data",
  tech = c("R"),
  img_src = "images/hex-CBEA.png"
)
```

```{r, eval=FALSE}
create_card(
  url = "https//github.com/qpmnguyen/microbesim"
  title = "microbesim: An R package for quick simulation of microbiome count data",
  source = "https://github.com/qpmnguyen/microbesim",
  description = "In order to develop and apply data analysis techniques to microbiome relative abundance data, researchers often have to perform numerical simulations that approximate real data. This R package was developed in order to collect together all the common ways to simulate microbiome relative abundance data and provide a uniform interface for quick-and-dirty simulations.",
  tech = c("R")
)
```

## Fun software  

```{r}
create_card(
  url = "https//github.com/qpmnguyen/nprbooksR",
  title = "nprbooksR: An R data package for the NPR's annual 'Books we love' recommendations",
  source = "https://github.com/qpmnguyen/nprbooksR",
  description = "Data from the NPR's 'Books we love' (formerly Book Concierge) recommendation list from 2013-2021",
  tech = c("R", "RSelenium")
)
```

## Shiny applications  

- [**DIFUSE Eddy Covariance Exploration**](https://github.com/diffuse-dartmouth/20x_BIO16): An RShiny app for BIOL 16 course at Dartmouth developed as part of the DIFUSE program (funded by NSF). This is an interactive data visualization and analysis tool for students to explore the eddy covariance techniques. Students are able to view data in raw format, play around with different visualizations, as well as being able to use such data to learn about how carbon exchange differs between seasons and time of day through building an interactive regression model.   

## Other contributions  

Things I helped write but not the main author/contributor  

- [**SBICgraph**](https://github.com/hoenlab/SBICgraph): Structural Bayesian Information Criterion (SBIC) for model selection in network models with a two-step algorithm to generate candidate model pool. ([*Zhou et al. 2021*](https://academic.oup.com/biostatistics/advance-article-abstract/doi/10.1093/biostatistics/kxab006/6171092?redirectedFrom=fulltext))

   
